---
title: "Intro to R programming and useful tips"
author: "Ryan James"
date: "2/24/2022"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = T, cache = T)
```

## Getting to know the basics 

R is a programming language that has become the standard in Ecology due to its flexibility and open source nature. R can be used from simple math to complex models and is very useful for generating figures. R, like all computer languages, using a specific syntax to run commands that it is programmed to do. In other words, R will only do what it is commanded to do, and therefore, many common errors are due to errors in syntax (e.g. misspellings, missed commas, or unclosed brackets). 

This example gives a basic intro into R syntax that can be useful for ecological research. This script gives examples of how to:

1.  Basic operations in R
2.  Assigning objects
3.  Types of data structures in R
4.  Functions in R
5.  Using Packages in R
      + How to install and load packages
6.  `tidyverse`
      + tidy data
      + piping 
7.  Working with `dataframes` and `tibbles`
      + Renaming and making columns
      + Summarizing data
      + Merging and combining mulitple dataframes
8.  Indexing
9.  Conditional Statements
      + Base R
      + Vectorized
      + `case_when`
10.  For loops
11.  Vector operations 
      + Custom Functions
      + `purr`
12.  Random sampling in R
13.  Figures with `ggplot2`
      + Combining plots

## Basic operations in R

R is useful for basic operations and follows math rules (i.e. PEMDAS). R will all code on a line unless there is a `#` to the left.

```{r math}
# addition 
1+1 

1+1 # + 2 (won't run anything to right of #)

# subtraction
5-2 

# multiplication
4*5

# division
33/5

# exponents can be done 2 ways
2^2
2**2

# follows PEMDAS
1+5*4
# different answer than above
(1+5)*4
```

Note the `[1]` appears next to your result. R is just letting you know that this line begins with the first value in your result. Some commands return more than one value, and their results may fill up multiple lines. For example, the command 100:130 returns 31 values; it creates a sequence of integers from 100 to 130. Notice that new bracketed numbers appear at the start of the first and second lines of output. These numbers just mean that the second line begins with that value. You can mostly ignore the numbers that appear in brackets:

```{r}
100:130
```
## Assigning objects
When working in R it is useful to store data as an object. Assigning objects can be done in multiple ways, but the most common are `<-` and `=`. These objects are stored in the R environment and can be called. Objects can be assigned multiple times, but only the last assignment is what is stored. Also it is important to know that R is case sensative and capital and lower case numbers are different.

```{r assign}
# assign an object
a = 4 
a

b <- 23

a+3 

b/2

a*b

c = 8
c = 14
c

d = 15 
D = 1 
d
D
```

## Types of data structures in R
R has 6 basic data types. (In addition to the five listed below, there is also raw which will not be discussed in this workshop.)

  + integer
  + numeric (real or decimal)
  + character
  + logical
  + complex
  
integers are whole numbers

numeric are numbers with decimals. Integers and numeric are different because of how the underlying data is stored. Other programming languages can use something similar as decimal, float, or double data types, which all slightly differ in how data is stored but are numbers that include decimals.

characters are strings of letters and numbers (e.g. `"abc"` and `"b1x"`) and are designated in R by `" "`. When using characters, `" "` are required because in R letters without quotations are objects and `c = 'd'` is different than `c = d`

logical is `TRUE` or `FALSE`. One thing to note is that `T` is the same as `TRUE` and `F` is the same as `FALSE`. Because `T` and `F` are special in R they cannot be used to name objects (but `t` and `f` are ok because R is case sensative). This is true for other cases as well like `NA` and `NULL`.  

complex numbers have both real and imaginary parts (`1+4i`)

Elements of these data types may be combined to form data structures, such as atomic vectors. When we call a vector atomic, we mean that the vector only holds data of a single data type. A vector is the most common and basic data structure in R and is pretty much the workhorse of R. Technically, vectors can be one of two types:
  + atomic vectors
  + lists
although the term “vector” most commonly refers to the atomic types not to lists. Lists differ because they can take on different data structures and can be more complex.

There are different ways to make vectors

```{r}

# make a numeric vector
a = c(1.1,5,3,4)
a

# make a integer vector
b = 1:15
b

# make a character vector 
c = c('a', 'b', 'c')
c 


```

Because characters can be both letters and numbers, numbers in a vector with letters are stored as a character. These cannot be used for math operations, but integers and numeric data types can be used for math. 

```{r, error=TRUE}
a = 4.4
a / 1 


b = 6L # L can be used to keep a numeric as an integer, R typically defaults to numeric
b*3

# character
c = '1'
c*4
```

Another common way to store data is in a dataframe or tibble (special type of dataframe from the `tidyverse` package we will see below). This is a collection of atomic vectors with the same length. 

```{r}
b = data.frame(c1 = c(1,2,3), c2 = c('a','b','c'))
b

```

## Functions in R
R comes with functions that are used to do tasks. Functions take arguments to complete a task. Functions have the general format `function(argument1 = , argument2,...)` The types of data used and output of the function is specific to that function. Below are just a few useful examples. 

```{r}
# summary statistics of sequence of numbers
a = c(1.1,5,3,4)
mean(a) #mean
median(a) #median
sd(a) #standard deviation
quantile(a, 0.5) # quantile at 0.5 (median)

# make a sequence of numbers
b = 1:15
b
c = seq(1,15,1) #more flexibility than :
c
seq(4,20,2)

# information about objects
d = c('a', 'b', 'c')
typeof(d) 
typeof(c)
length(d)

# dataframe/tibble specific functions
e = data.frame(c1 = c(1,2,3), c2 = c('a','b','c'))
names(e) # column names
nrow(e) # number of rows
length(e) # for dataframe number of columns
str(e)# structure of data
```

## Using Packages in R
R comes with a lot of base functions that are available for use when you open R, but this does not contain all of the functions useful to your tasks in R. Since R is open source, many R users have created Packages that contain functions that can be downloaded. Which includes the very common `tidyverse`.

### How to install and load packages
Packages can be downloaded from CRAN or from Github. To download directly from Github other packages are needed. 
```{r, eval=FALSE}
install.packages('tidyverse') #from cran
```

Once downloaded, packages can be loaded into the R environment with `library()` function. Packages have to be loaded each R session. In addition functions can be called directly from a package with `::` in the format of `packageName::function()`. 

```{r}
library(tidyverse)

```

## tidyverse
[tidyverse](https://www.tidyverse.org/) is a collection of packages that use similar syntax and are used for data science in R. Coding in tidyverse is typically easy to read and understand, and has useful functions that have been adopted into newer versions of base R (e.g. piping). Tibbles are the tidyverse version of a dataframe.
```{r}
c = tibble(c1 = c(1,2,3), c2 = c('a','b','c'))
c
```

### tidy data
Data is collected and stored in many different ways, which can make it difficult to analyze. One of the goals of tidyverse is to easily turn messy data into tidy data which can easily be analyzed. In tidy data:

1. Every column is a variable.
2. Every row is an observation.
3. Every cell is a single value.

Two functions `pivot_longer()` and `pivot_wider()` are useful in manipulating data stored in rows and columns. ***Note that `pivot_longer()` and `pivot_wider()` have replaced `gather()` and `spread()` in newer versions of `tidyverse`

```{r}
#tidying data 
stock = tibble(name = c('GOOG', 'AMC', 'GME'),
               Jan = c(1000, 2, 4),
               Feb = c(1010, 15, 30),
               March = c(1005, 25, 180))

df = stock %>% 
  pivot_longer(cols = Jan:March, 
               names_to = 'Month',
               values_to = 'Price')

df

# wide format
fish = tibble(species = rep(c('Salmon', 'Cod'),times = 3),
              year = rep(c(1999,2005,2020), each = 2),
              catch = c(50, 60, 40, 50, 60, 100))
fish 

fish %>%
  pivot_wider(id_cols = species, 
              names_from = year,
              values_from = catch)
```

### piping  
Tidyverse has an operator `%>%` known as a pipe that is useful for when you want to do multiple actions to the same data. It takes the output of the left of the `%>%` and makes it the first argument of what is on the right. Allowing to reduce code and make things tidier.

```{r}
# this code
df = as_tibble(mtcars)
df = filter(df, mpg > 20)
df = mutate(df, color = 'red')
df = select(df, mpg, cyl, color)

head(df)

# can become

df = mtcars %>%
  as_tibble()%>%
  filter(mpg > 20)%>%
  mutate(color = 'red')%>%
  select(mpg, cyl, color)

head(df)
```

## Working with `dataframes` and `tibbles`
Using either `dataframes` or `tibbles` will likely be the most common data structure for ecological data. Making these data structures is easy with the `data.frame()` or `tibble()` functions. Tibbles have more flexibility than dataframes and are part of the `tidyverse`. Dataframes are base R. When reading in tabular data, `read.csv()` will create a dataframe, while `read_csv()` will generate a tibble. 

```{r workDF}
# create a dataframe

df = data.frame(name = c('GOOG', 'AMC', 'GME'),
                Jan = c(1000, 2, 4),
                Feb = c(1010, 15, 30),
                March = c(1005, 25, 180))

df

# create a tibble
tib = tibble(name = c('GOOG', 'AMC', 'GME'),
             Jan = c(1000, 2, 4),
             Feb = c(1010, 15, 30),
             March = c(1005, 25, 180))

tib

# read in data file 
read.csv('data/Ov_CI.csv')

read_csv('data/Ov_CI.csv')
```

### Renaming and making columns
There are a few different ways to create a new column. The base R way is to use `$` with the object name of the dataframe on the left and the new column name on the right. This can be used to do vector operations as well. The other way is to the `mutate()` function which is part of the `dplyr` package in tidyverse. This function alows for more flexibility and can be very useful. The easiest way to rename columns is with `dplyr` functions like `rename()` or within function like `select()`.

```{r}
df = tibble(name = c('GOOG', 'AMC', 'GME'),
            Jan = c(1000, 2, 4),
            Feb = c(1010, 15, 30),
            March = c(1005, 25, 180))

df$new = 'new column'

df$tot = df$Jan + df$Feb + df$March

df 

# using mutate
df = df %>% 
  mutate(newCol = 'blue')

# multiple columns at a time
df = df %>%
  mutate(sum = Jan + Feb + March, 
         big = sum > 500)
df

df %>%
  rename(Name = name, January = Jan, February = Feb)

 

df %>%
  select(Name = name, January = Jan, sum, everything())
```

### Summarizing data
There are a few different useful ways to summarize the data in a dataframe or tibble. If you want to know everything about the dataframe, then the base function `summary()` is useful. If you would like to have more control to create summary tables, then `dplyr::summarize()` or `dplyr::summarise()` are great. This can be paired with `group_by()` to summarize over specific groups of data.

```{r}
summary(iris)

iris %>% 
  summarize(mean(Petal.Width),
            sd(Petal.Width))

iris %>% 
  group_by(Species)%>%
  summarize(mean(Petal.Width),
            sd(Petal.Width))
```

### Merging and combining mulitple dataframes
Combining data together is very common, and depending on the type of combination needed. 

#### Binding
If data has the same column names and needs to paste together, then `rbind()` and `dplyr::bind_rows()` are the tools need. For `rbind()`, the column names need to have the same name. `bind_rows()` does not have this problem.

```{r, error=T}
# bind data together 
sal = tibble(species = rep(c('Salmon'),times = 3),
              year = c(1999,2005,2020),
              catch = c(50, 60, 40))

cod = tibble(species = rep('Cod', times = 3),
             year = c(1999,2005,2020),
             catch = c(50, 60, 100))

crab = tibble(species = rep('Crab', times = 3),
              catch = c(50, 60, 100),
              effort = c(20, 30, 50))

rbind(sal,cod)

rbind(sal, crab)

bind_rows(sal, cod)

bind_rows(sal, crab)

```

#### Merge/Join
If two data frames contain different columns of data, then they can be merged together with the family of join functions.

 +`left_join()` = uses left df as template and joins all matching columns from right df 
 +`right_join()` = uses right df as template and joins all matching columns from left df
 +`inner_join()` = only matches columns contained in both dfs
 +`full_join()` = combines all rows in both dfs

```{r}
left = tibble(name = c('a', 'b', 'c'),
              n = c(1, 6, 7), 
              bio = c(100, 43, 57))

right = tibble(name = c('a', 'b', 'd', 'e'),
               cals = c(500, 450, 570, 600))

left_join(left, right, by = 'name')

right_join(left, right, by = 'name')

inner_join(left, right, by = 'name')

full_join(left, right, by = 'name')

# multiple matches
fish = tibble(species = rep(c('Salmon', 'Cod'),times = 3),
              year = rep(c(1999,2005,2020), each = 2),
              catch = c(50, 60, 40, 50, 60, 100))

col = tibble(species = c('Salmon', 'Cod'),
             coast = c('West', 'East'))

left_join(fish, col, by = 'species')


```

##  Indexing
Once data is stored in an object, being able to retrieve those values is useful. Referred to as indexing, the syntax is specific to how the data is stored. With indexing specific values within your object can be modified. 

```{r}
# vector 
b = 1:15
# 3rd object 
b[3]

# make a character vector 
c = c('a', 'b', 'c')
c
# 2nd object
c[2]
# change 
c[2] = 'new'
c

# dataframe and tibbles
mtcars
# first column
mtcars[1]
# first row
mtcars[1,]
# 2nd row of first column
mtcars[2,1]
# can call specific columns (called as a vector)
mtcars$mpg
mtcars$cyl
#same for tibble
d = mtcars %>% as_tibble
d[1]
d$mpg
d$cyl
# specific row in specific column
mtcars$cyl[1]
d$cyl[1]


```

## Conditional statements
In programing there are times that if something is true then you want an operation to occur, but not when a condition is not true. 
### Base R
These can be done with `if` and `if else` statements in base R. These are written if a condition is true then the operation is done. They can be built upon with `else if` if the first condition is false to do test a second condition. If you want it to be If true and if false do something else then `if` and `else` structure can be used. 

```{r}
b = 5 

if (b == 5){
  cat('b = 5 \n')
}

if (TRUE){
  c = 6
}else(
  c = 10
)
c 
if (F){
  c = 6
}else(
  c = 10
)
c 

if (b == 10){
  cat('b = 10 \n')
}else if (b == 5){
  cat('it worked \n')
}else{
  cat('nothing')
}

```
### `dplyr` functions
`dplyr` has two functions that are very useful for conditional statements. Because they are a function they can be vectorized which will be useful as you see below. `if_else()` is a function that based on if the input is `TRUE` or `FALSE` produces a different answer. `case_when()` is more flexible and allows for multple outputs based on conditions being `TRUE`

```{r}
x = 1:20

if_else(x > 10,
        'x > 10',
        'x < 10')


case_when(x < 6 ~ 'x < 6',
          between(x, 6, 15) ~ '6 < x < 15',
          x > 15 ~ 'x > 15')



```


##  For loops
Another useful tool in programming is `for` loops. For loops repeat a process for a certain number of iterations. These can be useful iterate over a dataset or when using information in a time series. The `for` loop works over the number sequence indicated and does the code within the loop (inside of `{}`) for each number in the sequence. The iteration is typically indicated with `i`, but is just an object that is replaced at the begining of each loop and can be anything.

```{r}
for(i in 1:10){
  print(i)
}

for(turtle in 5:10){
  print(turtle)
}

for(flower in 1:nrow(iris)){
  cat('The species for this iteration is ',
      as.character(iris$Species[flower]), '\n')
}

d = seq(1,15, 2)
d
for(i in 1:length(d)){
  b = d[i] + 1
  cat('d =',d[i], 'b = d + 1 =', b, '\n' )
}

b = 1:10
for (i in 2:10){
  z = b[i] - b[i-1]
  
  cat('z =', z, 'b[i] =', b[i], 'b[i-1] =', b[i-1], '\n')
}


start = 10 
pop = tibble(time = 0:10, n = NA)
pop$n[pop$time == 0] = start
pop
for (t in 1:10){
  growth = rnorm(n =1, mean = 3, sd = 1)
  pop$n[pop$time == t] = growth + pop$n[pop$time == (t-1)]
}
pop


start = 1000 
pop = tibble(time = 0:10, n = NA)
pop$n[pop$time == 0] = start
pop
for (t in 1:10){
  growth = rnorm(n =1, mean = 300, sd = 50)
  death = rnorm(n = 1, mean = 300, sd = 50)
  pop$n[pop$time == t] = growth - death + pop$n[pop$time == (t-1)]
}
pop


```
##  Vector operations 
As we have seen above, we can do operations over vectors. We sometimes want to do this to vectors stored in dataframes/tibbles, and the `mutate()` function makes this easy. 

```{r}
iris %>% 
  mutate(petalArea = Petal.Length*Petal.Width)

iris %>%
  mutate(petalArea = Petal.Length*Petal.Width,
         PetalSize = if_else(condition = petalArea > 0.2, true ='big',
                             false = 'small'))

iris %>%
  mutate(petalArea = Petal.Length*Petal.Width,
         PetalSize = if_else(condition = petalArea > 0.2, true ='big',
                             false = 'small'))%>%
  group_by(PetalSize)%>%
  summarize(mean = mean(Petal.Width),
            n())


```
### Custom functions
So far we have used functions that are apart of base R or from different packages, but we can also build custom functions. Custom functions are useful when existing functions will not do the task at hand, or when combining functions over multiple times. These can be used for vector operations.

```{r}
# custom funciton 
my_fx = function(x){
  b = x * 23
  return(b)
}

my_fx(5)

t = seq(2,10,2)
my_fx(t)

# custom function with multiple arguments
my_fx2 = function(x,y){
  b = x/y
  return(b)
}

my_fx2(4,5)

# custom fucntion with true false

my_fx3 = function(x = T){
  if (x == T){
    cat('x is true \n')
  }else{
    cat('x is not true \n')
  }
}

my_fx3(x = T)
my_fx3()
my_fx3(x = F)
my_fx3(x = 5)


# custom functions with vector operations
mi_km = function(mi){
  km = mi * 1.60934
  return(km)
}

mtcars %>%
  mutate(kmpg = mi_km(mpg))

```

### `purr`
The newest and new standard package with `tidyverse` is `purr` with its set of `map()` functions. Some similarity to `plyr` (and base) and `dplyr` functions but with more consistent names and arguments. Notice that map function can have some specification for the type of output.
  + `map()` makes a list.
  + `map_lgl()` makes a logical vector.
  + `map_int()` makes an integer vector.
  + `map_dbl()` makes a double vector.
  + `map_chr()` makes a character vector.

```{r}

df = iris %>%
  select(-Species)
#summary statistics
map_dbl(df, mean)
map_dbl(df, median)
map_dbl(df, sd)

###
#Creating models by group
#Similar of what we did with plyr and dplyr examples
###
data(mtcars)

models <- mtcars %>% 
  split(.$cyl) %>% 
  map(~lm(mpg ~ wt, data = .))

models %>% 
  map(summary) %>% 
  map_dbl(~.$r.squared) #or map_dbl("r.squared")

###
#Mapping over multiple arguments
###

mu <- list(5, 10, -3)
mu %>% 
  map(rnorm, n = 5) %>% #rnorm - function to extract values from a normal continuous distribution based on some parameters
  
str()
#> List of 3

#Adding SD as well
sigma <- list(1, 5, 10)
seq_along(mu) %>% 
  map(~rnorm(5, mu[[.]], sigma[[.]])) %>% 
  str()
#> List of 3

###
#map2 or pmap allows you to iterate over two or more vectors in parallel
###
map2(mu, sigma, rnorm, n = 5) %>% str()

#or with pmap
n <- list(1, 3, 5)
args2 <- list(mean = mu, sd = sigma, n = n)
args2 %>% 
  pmap(rnorm) %>% 
  str()

```

## Random sampling in R
Generating random numbers or randomly sampling is commonly required in ecology. R has base functions that can be used to randomly generate numbers depending on the distribution wanted. Also packages like [truncnorm](https://www.rdocumentation.org/packages/truncnorm/versions/1.0-8/topics/truncnorm) and  `dplyr` in `tidyverse` are useful.  

```{r, results='hide'}
library(truncnorm)
library(tidyverse)
```


#### Generate random numbers
```{r}
# base R
rnorm(5, mean = 0, sd = 1)
runif(5, min = 0, max = 5)
rbinom(5, size = 2, prob = 0.5)
rbinom(5, size = 2, prob = 0.2)

```

```{r}
#library(rtruncnorm)
# truncnorm to truncate normal distribution
rnorm(20, mean = 1, sd = 2)
truncnorm::rtruncnorm(20, mean = 1, sd = 2, a=-Inf, b=Inf)
truncnorm::rtruncnorm(20, mean = 1, sd = 2, a=0, b=2)
```

#### Random selection 
```{r}
# base R 
b = 1:10
sample(b, size =2, replace = F)

c = 1:5 
sample(c, size = 6, replace = T)

# from list
sample(c('good', 'bad'), size = 8, replace = T)

# change probability
sample(c('good', 'bad'), size = 8, replace = T, prob = c(0.2, 0.8))

# sample matrices from list 
m1 = matrix(c(6,5,4,3,2,1), ncol = 2)
m2 = matrix(c(1,2,3,4,5,6), ncol = 2)
m3 = matrix(c(1,2,3,4,5,6), ncol = 2, byrow = T)

m = list(m1, m2, m3)

sample(m, size = 4, replace = T)

sample(m, size = 4, replace = T, prob = c(0.8, 0.1, 0.1))

```


```{r}
#library(tidyverse)
df = tibble(cond = c('good', 'bad', 'ok'), prob = c(0.5, 0.3, 0.2))

dplyr::sample_n(df, size = 2, replace = F)

dplyr::sample_n(df, size = 3, replace = T)

dplyr::sample_n(df, size = 10, replace = T, weight = prob)
```

We can bring what we learned together to simulate change in a population over time. This code simulates the change in a population over `time_length` with a starting population size `pop_start` using a `for` loop. The code is replicated `reps` times. The population size changes by randomly generating births and deaths from a table of with various types of years `year_type`.


```{r}
# table of year conditions
year_type = tibble(cond = c('good', 'average', 'bad'),
              birth = c(500, 300, 100),
              death = c(100, 300, 500),
              prob = c(0.25, 0.5, 0.25))

# beginning size of population
pop_start = 50000 
# number of time iterations to use
time_length = 100
# number of replicate trials to simulate
reps = 10

# for loop for number of reps
for (i in 1:reps){
  # indicate which trial based on i in 1:reps
  trial = paste0('trial_',i)
  # generate a new table to store output with rows 0:time_length for given trial
  pop = tibble(time = 0:time_length, n = NA,
               trial = trial, year = NA, change = NA)
  # for row where time == 0 (initial conditions) make n = to starting population
  pop$n[pop$time == 0] = pop_start
  
  # for loop over all time points 
  for (t in 1:(nrow(pop)-1)){
    # randomly select year conditions from year_type df that is weighted by the probability of occurance
    year = sample_n(year_type, size = 1, weight = prob)
    # randomly generate number of births based on year type conditions 
    birth = round(rnorm(n = 1, mean = year$birth[1], sd = 50))
    # randomly generate number of deaths based on year type conditions 
    death = round(rnorm(n = 1, mean = year$death[1], sd = 50))
    # population size (n) at time t is calculated by adding the difference in births and deaths to the population size where time = t - 1
    pop$n[pop$time == t] = pop$n[pop$time == (t-1)] + birth - death 
    # label year type for time = t
    pop$year[pop$time == t] = year$cond[1]
    # cacluate change in population size at time t
    pop$change[pop$time == t] = birth - death
  }
  
  # this stores the output of each trial in a new tibble (tpop) if it is the first trial then pop is tpop, after first trial then bind the rows of the trial to the rest of the trials 
  if (i ==1){
    tpop = pop
  }else{
    tpop = bind_rows(tpop,pop)
  }
}

head(tpop)
```


## Figures with `ggplot2` 
The `ggplot2` package is part of the packages that load with `tidyverse` and has become the standard in ecology. The syntax builds upon on a base function and is very customizable [see cheat sheet](https://www.rstudio.com/resources/cheatsheets/). 

The base of all `ggplot2` begins with `ggplot()` and `geom_...()` are built upon them 

```{r plot}
ggplot(tpop, aes(x = time, y = n))+
  geom_point()
```

Show color based on trial number and line connecting dots

```{r}
ggplot(tpop, aes(x = time, y = n, color = trial))+
  geom_point()+
  geom_line()
```

Change labels and style of plot


```{r}
ggplot(tpop, aes(x = time, y = n, color = trial))+
  geom_point()+
  geom_line()+
  labs(x = 'Time', y = 'Population size')+
  theme_classic()
```

Modify the size of axis label text and legend position  

```{r}
ggplot(tpop, aes(x = time, y = n, color = trial))+
  geom_point()+
  geom_line()+
  labs(x = 'Time', y = 'Population size')+
  theme_classic()+
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.position = 'bottom',
        legend.title = element_blank())
```

Modify the range of time on x axis

```{r}
ggplot(tpop, aes(x = time, y = n, color = trial))+
  geom_point()+
  geom_line()+
  scale_x_continuous(limits = c(0,20))+
  labs(x = 'Time', y = 'Population size')+
  theme_classic()+
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.position = 'bottom',
        legend.title = element_blank())
```

Split each trial into own grid

```{r}
ggplot(tpop, aes(x = time, y = n))+
  geom_point()+
  geom_line()+
  labs(x = 'Time', y = 'Population size')+
  facet_wrap(~trial)+
  theme_classic()+
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.position = 'bottom',
        legend.title = element_blank())
```

Modify the label and size of strip text


```{r}
labels = c('trial_1' = 'Trial 1',
          'trial_2' = 'Trial 2',
          'trial_3' = 'Trial 3',
          'trial_4' = 'Trial 4',
          'trial_5' = 'Trial 5',
          'trial_6' = 'Trial 6',
          'trial_7' = 'Trial 7',
          'trial_8' = 'Trial 8',
          'trial_9' = 'Trial 9',
          'trial_10' = 'Trial 10')

ggplot(tpop, aes(x = time, y = n))+
  geom_point()+
  geom_line()+
  labs(x = 'Time', y = 'Population size')+
  facet_wrap(~trial, nrow = 2, labeller = as_labeller(labels))+
  theme_classic()+
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.position = 'bottom',
        legend.title = element_blank(),
        strip.text = element_text(size = 12))
```



Remake figure with the mean and min and max values from the multiple trials


```{r}
ggplot(tpop, aes(x = time, y = n))+
  geom_pointrange(stat = "summary",
                  fun.min = 'min',
                  fun.max = 'max',
                  fun = 'mean')+
  labs(x = 'Time', y = 'Population size')+
  theme_classic()+
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12))
```

Make box plot of change in population size based on type of year 

```{r}
ggplot(tpop, aes(x = year, y = change))+
  geom_boxplot()+
  labs(x = 'Type of year', y = expression(Delta*'Population'))+
  theme_bw()
```

Remove `NA` and add color to plot

```{r}
ggplot(tpop %>% drop_na, aes(x = year, y = change, fill = year))+
  geom_boxplot()+
  labs(x = 'Type of year', y = expression(Delta*'Population'))+
  theme_bw()
```

Change order of x axis and color of plot. Colors can be both hex code or from names that R has. A help website for picking colors is [here](https://rstudio-pubs-static.s3.amazonaws.com/3486_79191ad32cf74955b4502b8530aad627.html).

```{r ch}

tpop = tpop %>% 
  mutate(year = factor(year, levels = c('good', 'average', 'bad')))%>%
  drop_na

colors = c('good' = 'darkgreen',
           'average' = 'goldenrod',
           'bad' = 'firebrick')

ggplot(tpop, aes(x = year, y = change, fill = year))+
  geom_boxplot()+
  labs(x = 'Type of year', y = expression(Delta*'Population'))+
  scale_fill_manual(values = colors)+
  theme_bw()
```

Modify the labels and remove the legend

```{r}
ggplot(tpop, aes(x = year, y = change, fill = year))+
  geom_boxplot()+
  labs(x = 'Type of year', y = expression(Delta*'Population'))+
  scale_x_discrete(labels = c('Good', 'Average', 'Bad'))+
  scale_fill_manual(values = colors)+
  theme_bw()+
  theme(axis.title = element_text(size = 18), 
        axis.text.y = element_text(size = 18, colour = "black"), 
        axis.text.x = element_text(size = 14, colour = "black"), 
        legend.position = 'none',
        legend.title = element_blank(),
        strip.text.x = element_text(size = 18),
        legend.text = element_text(size = 12))
```

### Combining plots 
Sometimes we would like to combine different sub figures together to make a single figure. There are a few packages that can do this with `ggpubr` and `patchwork` some of the most common. I like `ggpubr` and use this one, but people seem to like `patchwork`. 

```{r}
library(ggpubr)
library(wesanderson)

a = ggplot(tpop, aes(x = time, y = n))+
  geom_pointrange(stat = "summary",
                  fun.min = 'min',
                  fun.max = 'max',
                  fun = 'mean')+
  labs(x = 'Time', y = 'Population size')+
  theme_classic()+
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12))

b = ggplot(tpop, aes(x = year, y = change, fill = year))+
  geom_boxplot()+
  labs(x = 'Type of year', y = expression(Delta*'Population'))+
  scale_x_discrete(labels = c('Good', 'Average', 'Bad'))+
  scale_fill_manual(values = colors)+
  theme_bw()+
  theme(axis.title = element_text(size = 18), 
        axis.text.y = element_text(size = 18, colour = "black"), 
        axis.text.x = element_text(size = 14, colour = "black"), 
        legend.position = 'none',
        legend.title = element_blank(),
        strip.text.x = element_text(size = 18),
        legend.text = element_text(size = 12))

ggarrange(a,b,
          labels = c('a)','b)'),
          ncol = 1)

ggarrange(a,b,
          labels = c('a)','b)'),
          ncol = 1,
          align = 'hv')

# common legend
a = ggplot(mtcars, aes(wt, fill = as.character(cyl), 
                       color = as.character(cyl)))+
  geom_density(alpha = 0.4)+
  labs(x = 'Weight of car (tonnes)', 
       fill = '# of engine cylinders')+
  scale_color_manual(values = wes_palette('GrandBudapest1'),
                     guide = "none")+
  scale_fill_manual(values = wes_palette('GrandBudapest1'))+
  theme_bw()+
  theme(axis.title = element_text(size = 10), 
        axis.text.y = element_text(size = 10, colour = "black"), 
        axis.text.x = element_text(size = 8, colour = "black"), 
        legend.position = 'bottom',
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text.x = element_text(size = 10),
        legend.text = element_text(size = 7))

b = ggplot(mtcars, aes(mpg, color = as.character(cyl),
                       fill = as.character(cyl)))+
  geom_density(alpha = 0.4)+
  labs(x = 'Miles/gallon',
       fill = '# of engine cylinders')+
  scale_color_manual(values = wes_palette('GrandBudapest1'),
                     guide = "none")+
  scale_fill_manual(values = wes_palette('GrandBudapest1'))+
  theme_bw()+
  theme(axis.title = element_text(size = 10), 
        axis.text.y = element_text(size = 10, colour = "black"), 
        axis.text.x = element_text(size = 8, colour = "black"), 
        legend.position = 'bottom',
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text.x = element_text(size = 10),
        legend.text = element_text(size = 7))
  

c = ggplot(mtcars, aes(wt, mpg, group = cyl, color = as.character(cyl)))+
  geom_point(size = 2)+
  geom_smooth(method = 'lm',size = 1)+
  labs(x = 'Weight of car (tonnes)', 
       y = 'Miles/gallon',
       color = '# of engine cylinders')+
  scale_color_manual(values = wes_palette('GrandBudapest1'))+
  theme_bw()+
  theme(axis.title = element_text(size = 10), 
        axis.text.y = element_text(size = 10, colour = "black"), 
        axis.text.x = element_text(size = 8, colour = "black"), 
        legend.position = 'bottom',
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text.x = element_text(size = 10),
        legend.text = element_text(size = 7))


ggarrange(a,b,c, 
          labels = c('A','B','C'),
          nrow = 2,ncol = 2,
          common.legend = F)

ggarrange(a,b,c, 
          labels = c('A','B','C'),
          nrow = 2, ncol = 2,
          common.legend = T,
          legend = 'top')

ggarrange(ggarrange(a,b, labels = c('A','B'), common.legend = T),c,
          labels = c('','C'),
          nrow = 2,
          legend = 'none')
  
```

